---
title: "Assignment 3: ML Exercise – Multiple Regression with TEDS 2016"
author: "EPPS 6323 Knowledge Mining"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-tools: true
    self-contained: true
execute:
  warning: false
  message: false
---

## Overview

This assignment uses the **Taiwan Election and Democratization Study (TEDS) 2016** dataset to explore multiple regression, build a custom plotting function, and critically evaluate the limits of OLS when the dependent variable is binary.

------------------------------------------------------------------------

## 1. Setup and Data Loading

```{r}
#| label: setup
#| message: false

library(haven)
library(ggplot2)
library(dplyr)

# Load the TEDS 2016 dataset
TEDS_2016 <- read_stata(
  "https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true"
)
```

------------------------------------------------------------------------

## 2. Exploring the Dataset

```{r}
#| label: explore

# Quick overview
dim(TEDS_2016)
names(TEDS_2016)

# Key variables
str(TEDS_2016[, c("votetsai", "age", "edu", "income")])

# Distribution of the dependent variable
table(TEDS_2016$votetsai, useNA = "always")
```

The **dependent variable** is `votetsai` — whether a respondent voted for Tsai Ing-wen in the 2016 presidential election.

------------------------------------------------------------------------

## 3. Multiple Regression Model

We run a multiple regression model predicting `votetsai` from `age`, `edu` (education), and `income`.

```{r}
#| label: multiple-regression

model <- lm(votetsai ~ age + edu + income, data = TEDS_2016, na.action = na.omit)
summary(model)
```

**Interpretation:** The OLS model estimates the linear relationship between each predictor and the probability of voting for Tsai. However, as we will see below, using OLS with a binary outcome has fundamental problems.

------------------------------------------------------------------------

## 4. The `regplot` Function

We define a custom function that combines `lm()`, `plot()`, and `abline()` to produce a simple regression fit plot.

```{r}
#| label: regplot-function

## Creating a function: regplot
## Combines lm, plot, and abline to create a regression fit plot

regplot <- function(x, y, xlab = "Predictor", ylab = "votetsai",
                    main = "Regression Fit") {
  fit <- lm(y ~ x)
  plot(x, y,
    xlab  = xlab,
    ylab  = ylab,
    main  = main,
    pch   = 16,
    col   = rgb(0.2, 0.4, 0.8, 0.3),
    cex   = 0.7
  )
  abline(fit, col = "red", lwd = 2)
}
```

------------------------------------------------------------------------

## 5. Regression Plots

### 5a. `votetsai` \~ Age

```{r}
#| label: regplot-age
#| fig-cap: "Linear regression of votetsai on Age"

regplot(
  x    = TEDS_2016$age,
  y    = TEDS_2016$votetsai,
  xlab = "Age",
  main = "votetsai ~ Age"
)
```

### 5b. `votetsai` \~ Education

```{r}
#| label: regplot-edu
#| fig-cap: "Linear regression of votetsai on Education"

regplot(
  x    = TEDS_2016$edu,
  y    = TEDS_2016$votetsai,
  xlab = "Education",
  main = "votetsai ~ Education"
)
```

### 5c. `votetsai` \~ Income

```{r}
#| label: regplot-income
#| fig-cap: "Linear regression of votetsai on Income"

regplot(
  x    = TEDS_2016$income,
  y    = TEDS_2016$votetsai,
  xlab = "Income",
  main = "votetsai ~ Income"
)
```

------------------------------------------------------------------------

## 6. What Is the Problem? Why?

### The Core Issue: Binary Dependent Variable

The dependent variable `votetsai` takes only **two values: 0 and 1**:

-   **0** = Did *not* vote for Tsai Ing-wen\
-   **1** = Did vote for Tsai Ing-wen

This creates **three critical problems** when using Ordinary Least Squares (OLS):

| Problem | Explanation |
|------------------------------------|------------------------------------|
| **Out-of-range predictions** | OLS can produce predicted values below 0 or above 1, which are nonsensical as probabilities |
| **Non-linear relationship** | The true relationship between predictors and a binary outcome is inherently S-shaped (sigmoid), not linear |
| **Heteroskedasticity** | With a binary outcome, the variance of errors depends on the predicted value — a core OLS assumption is violated |

The scatter plots make this visually clear: all the data points lie on **two horizontal bands** (y = 0 and y = 1), and the red regression line cuts awkwardly through them. No amount of tuning the OLS model will fix this structural mismatch.

> **In short:** OLS was designed for continuous outcomes. Forcing it onto a binary outcome produces inefficient estimates and uninterpretable predictions.

------------------------------------------------------------------------

## 7. How to Improve Prediction of the Dependent Variable

### Recommended Approach: Logistic Regression

Since `votetsai` is binary, the appropriate model is **logistic regression** (a Generalized Linear Model with a binomial family and logit link). It models the **log-odds** of voting for Tsai, keeping all predicted probabilities between 0 and 1.

```{r}
#| label: logistic-regression

logit_model <- glm(
  votetsai ~ age + edu + income,
  data   = TEDS_2016,
  family = binomial(link = "logit"),
  na.action = na.omit
)

summary(logit_model)
```

### Model Comparison

```{r}
#| label: compare-models

# Predicted probabilities from logistic regression
pred_probs <- predict(logit_model, type = "response")

# Classification accuracy using 0.5 threshold
predicted_class <- ifelse(pred_probs >= 0.5, 1, 0)
actual <- as.numeric(TEDS_2016$votetsai[!is.na(TEDS_2016$votetsai) &
                                         !is.na(TEDS_2016$age) &
                                         !is.na(TEDS_2016$edu) &
                                         !is.na(TEDS_2016$income)])

accuracy <- mean(predicted_class == actual, na.rm = TRUE)
cat("Logistic Regression Accuracy:", round(accuracy * 100, 1), "%\n")

# Confusion matrix
table(Predicted = predicted_class, Actual = actual)
```

### Visualizing Logistic Fit

```{r}
#| label: logit-plot
#| fig-cap: "Logistic regression fit of votetsai on Education"

# Use education as the focal predictor for visualization
edu_range <- seq(min(TEDS_2016$edu, na.rm = TRUE),
                 max(TEDS_2016$edu, na.rm = TRUE), length.out = 200)

# Predict with mean age and income held constant
newdata <- data.frame(
  edu    = edu_range,
  age    = mean(TEDS_2016$age,    na.rm = TRUE),
  income = mean(TEDS_2016$income, na.rm = TRUE)
)
newdata$prob <- predict(logit_model, newdata = newdata, type = "response")

plot(TEDS_2016$edu, TEDS_2016$votetsai,
     xlab = "Education",
     ylab = "Pr(votetsai = 1)",
     main = "Logistic Regression: votetsai ~ Education\n(Age & Income held at mean)",
     pch  = 16,
     col  = rgb(0.2, 0.4, 0.8, 0.2),
     cex  = 0.7)

lines(newdata$edu, newdata$prob, col = "red", lwd = 2.5)
abline(h = 0.5, col = "gray50", lty = 2)
legend("topleft",
       legend = c("Logistic curve", "Decision boundary (0.5)"),
       col    = c("red", "gray50"),
       lty    = c(1, 2), lwd = 2, bty = "n")
```

### Summary of Improvements

| Approach | Why it helps |
|------------------------------------|------------------------------------|
| **Logistic Regression** | Constrains predictions to \[0,1\]; models non-linear probability |
| **Add more predictors** | Include party ID, political attitudes (e.g., `tomiddle`, `unification`) to improve fit |
| **Interaction terms** | e.g., `age × education` may capture nuanced effects |
| **Regularization (LASSO/Ridge)** | Handles multicollinearity if many predictors are added |
| **Tree-based models** | Random Forest or boosting can capture complex non-linearities |

------------------------------------------------------------------------

## Summary

-   OLS regression with a **binary dependent variable** is inappropriate — it violates fundamental assumptions and produces out-of-range predictions.
-   **Logistic regression** is the correct baseline model for binary outcomes.
-   The TEDS 2016 data shows that education is a particularly meaningful predictor of voting for Tsai Ing-wen, consistent with the literature on Taiwan's electoral politics.
