---
title: "Assignment 1: Brush up R and Quarto"
format: html
toc: true
---

## Overview

For this assignment, I reviewed basic R programming, practiced Quarto, and completed an exploratory data analysis using the TEDS_2016 dataset.

## Lab Files

-   [Lab01](Lab01.html)
-   [Lab02](Lab02.html)

## Load the Dataset

```{r}
library(haven)

TEDS_2016 <- haven::read_dta("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")
```

### Initial Inspection

```{r}
# Check dimensions of the dataset
dim(TEDS_2016)

# Summarize only the variables needed for the assignment
summary(TEDS_2016[c("Tondu", "votetsai", "female", "DPP", "age", "income", "edu", "Taiwanese", "Econ_worse")])
```

### Data Problems and Handling Missing Values

**4. What problems do you encounter when working with the dataset?**

After review of the initial summary statistics, I can identify two primary data integrity issues:

1.  **Incorrect Data Types:** The `haven` package imports several categorical variables (such as `Tondu`, `female`, `DPP`, `edu`, and `income`) as continuous numeric vectors. At the same time, R calculated meaningless descriptive statistics, like means and medians, for those categories. Because R thinks these categories are just math numbers, I need to convert them into proper groups before we can build any meaningful charts.
2.  **Missing Data:** The summary shows a substantial number of missing values (`NA`s). Most notably, the `votetsai` variable is missing 429 observations, and the `edu` variable is missing 10.

**5. How to deal with missing values?**

I have to be careful about how I handle missing data instead of just deleting it all at once.

It might be tempting to use a quick command like na.omit() to wipe out any row that has a missing value. However, that could be a bad idea for this dataset. The votetsai column alone is missing 429 values. If I remove every row that has a blank spot, I would throw away about 25% of the entire dataset which would shrink my sample size too much and likely skew the results.

A much better approach is to keep the data and just ignore the blank spots on a case-by-case basis. For example, when running calculations, na.rm = TRUE could be used to tell R to simply skip the missing numbers. When making charts, filters(!is.na(variable)) could be used to temporarily hide the missing data just for that specific graph. This way, I only ignore the blanks without throwing out good data from other columns.

### 6. Explore Relationships with Tondu

```{r}
# Convert the remaining binary variables into factors for accurate plotting
TEDS_2016$female <- factor(TEDS_2016$female, labels = c("Male", "Female"))
TEDS_2016$DPP <- factor(TEDS_2016$DPP, labels = c("Non-DPP", "DPP"))
TEDS_2016$Taiwanese <- factor(TEDS_2016$Taiwanese, labels = c("Not Taiwanese", "Taiwanese"))
TEDS_2016$Econ_worse <- factor(TEDS_2016$Econ_worse, labels = c("Econ Not Worse", "Econ Worse"))

# --- Part 1: Continuous/Ordinal Variables vs. Tondu ---
# Set up a 1x3 grid to plot Age, Income, and Education side-by-side
par(mfrow = c(1, 3)) 

boxplot(age ~ Tondu, data = TEDS_2016, main = "Age by Tondu", col = "lightgray", las = 2)
boxplot(income ~ Tondu, data = TEDS_2016, main = "Income by Tondu", col = "lightgray", las = 2)
boxplot(edu ~ Tondu, data = TEDS_2016, main = "Education by Tondu", col = "lightgray", las = 2)

# --- Part 2: Categorical Variables vs. Tondu ---
# Set up a 2x2 grid to plot the binary variables
# Note: R automatically generates spineplots when plotting two factors against each other
par(mfrow = c(2, 2))

plot(Tondu ~ female, data = TEDS_2016, main = "Tondu by Gender", col = c("coral", "lightblue", "lightgreen", "gold", "plum", "cyan", "gray"), las = 2)
plot(Tondu ~ DPP, data = TEDS_2016, main = "Tondu by DPP Support", col = c("coral", "lightblue", "lightgreen", "gold", "plum", "cyan", "gray"), las = 2)
plot(Tondu ~ Taiwanese, data = TEDS_2016, main = "Tondu by Taiwanese ID", col = c("coral", "lightblue", "lightgreen", "gold", "plum", "cyan", "gray"), las = 2)
plot(Tondu ~ Econ_worse, data = TEDS_2016, main = "Tondu by Econ Worse", col = c("coral", "lightblue", "lightgreen", "gold", "plum", "cyan", "gray"), las = 2)

# Reset the plot grid back to default (1x1) for any future charts
par(mfrow = c(1, 1))
```

### 7. What methods would you use?

The methods I chose depend entirely on the type of data being compared to `Tondu`. Since `Tondu` is a categorical variable broken into seven distinct groups, I can't just run a standard numeric correlation. Instead, I split my approach based on the variables:

-   **For continuous variables (`age`, `income`, `edu`):** I used boxplots. This makes it easy to visually compare the median and the spread of the data across the different `Tondu` preference groups.
-   **For categorical variables (`female`, `DPP`, `Taiwanese`, `Econ_worse`):** I used spineplots (which R generates by default when asked to plot two factors against each other). This lets me easily see the proportions of `Tondu` preferences within each specific group.

### 8. How about the votetsai variable?

The `votetsai` variable tracks whether the respondent voted for the DPP candidate, Tsai Ing-wen. Just like the other binary variables, I have to convert it into a factor first so R knows it is a category.

As I found during the initial data inspection, this variable is missing 429 values. But the good news is that by plotting it as a factor against `Tondu`, R will automatically ignore those missing (`NA`) values just for this specific chart, meaning I don't have to remove any rows from our overall dataset.

```{r}
# Convert votetsai to a factor with distinct labels
TEDS_2016$votetsai <- factor(TEDS_2016$votetsai, labels = c("Did not vote for Tsai", "Voted for Tsai"))

# Plot Tondu by votetsai
plot(Tondu ~ votetsai, data = TEDS_2016, 
     main = "Tondu Preference by Tsai Ing-wen Voters", 
     col = c("coral", "lightblue", "lightgreen", "gold", "plum", "cyan", "gray"), 
     las = 2)
```

### 9. Frequency Table and Barchart of Tondu

Finally, I need to generate a frequency table and a barchart for the main variable, `Tondu`. Before I can do that, I must assign the proper text labels to the numeric codes provided in the dataset.

While the assignment instructions suggest using `as.numeric()`, standard R graphics require the variable to be converted into a `factor` to properly display text labels on a barchart.

```{r}
# Assign categorical labels to the Tondu variable
TEDS_2016$Tondu <- factor(TEDS_2016$Tondu, 
                          levels = c(1, 2, 3, 4, 5, 6, 9),
                          labels = c("Unification now", 
                                     "Status quo, unif. in future", 
                                     "Status quo, decide later", 
                                     "Status quo forever", 
                                     "Status quo, indep. in future", 
                                     "Independence now", 
                                     "No response"))

# Convert the table to a data frame and format it clearly for HTML
tondu_freq <- as.data.frame(table(TEDS_2016$Tondu))
knitr::kable(tondu_freq, col.names = c("Tondu Preference", "Frequency"))

# Generate the barchart
barplot(table(TEDS_2016$Tondu), 
        main = "Distribution of Tondu Preferences",
        ylab = "Frequency",
        col = "steelblue",
        las = 2,             # Rotates x-axis labels to be vertical
        cex.names = 0.6)     # Shrinks label text so it doesn't get cut off
```
